---
title: "HPB_preprocess"
output: html_document
date: "2026-04-06"
---

!!!!! NOTE !!!!
Most of the HPB map processing and making was done in the RHESSys Tutorial project; this script is copied to show the general workflow, maps made in GRASS and what was input to GRASS are in the spatial/data folder, but the GRASS project and orignial making CCR_Bathy_NED are in the Tutorial repo.
The Bathy_NED file was just appened bathymetry to the DEM, so for HPB this makes functionally no difference from using the DEM that is downloaded in the GetDEM.R script.

Main change in this script is the addition and processing of the NLCD data to match the current maps and make new worldfiles

## Get packages 

```{r}
library(terra)
library(RHESSysPreprocessing)
```


##reproject DEM to utm for GRASS

```{r}
getwd()
ned_bathy <- rast("../CCR_NEW/Spatial_Data_Comp/Create_Maps/CCR_bathy/CCR_bathy_NED.tif")
plot(ned_bathy)
crs(ned_bathy)
#DEM is NAD83 EPSG4269

#convert to NAD83 UTM17N
dem_reproj <- project(ned_bathy, "EPSG:26917")
plot(dem_reproj)
crs(dem_reproj)

writeRaster(dem_reproj, "spatial_data/NAD83UTM17N_epsg26917/NED_bathy.tif", overwrite = T)


####watershed
ws <- rast("./spatial_data/NAD83_epsg4269/hpb_ws.tif")
plot(ws)

ws_reproj <- project(ws, "EPSG:26917")
plot(ws_reproj)
crs(ws_reproj)

writeRaster(ws_reproj, "spatial_data/NAD83UTM17N_epsg26917/hpb_ws.tif", overwrite = T)



```




#### look at maps 
sum(is.na(values(ws))) is to check for NAs, if value doesn't equal basins will need to fill those gaps 

```{r}
#basin
getwd()
ws <- rast("../spatial_data/NAD83UTM17N_epsg26917/hpb_ws.tif")
plot(ws, main = "Watershed")
sum(is.na(values(ws)))

#dem
dem <- rast("../spatial_data/NAD83UTM17N_epsg26917/GRASS_output/hpb_dem.tif")
plot(dem, main = "DEM")
sum(is.na(values(dem)))


#slope
slope <- rast("../spatial_data/NAD83UTM17N_epsg26917/GRASS_output/hpb_slope.tif")
plot(slope, main = "slope")
sum(is.na(values(slope)))


#aspect
aspect <- rast("../spatial_data/NAD83UTM17N_epsg26917/GRASS_output/hpb_aspect.tif")
plot(aspect, main = "aspect")
sum(is.na(values(aspect)))


#West Horizon
whr <- rast("../spatial_data/NAD83UTM17N_epsg26917/GRASS_output/hpb_whr.tif")
plot(whr, main = "West horizon")
sum(is.na(values(whr)))


#East horizon
ehr <- rast("../spatial_data/NAD83UTM17N_epsg26917/GRASS_output/hpb_ehr.tif")
plot(ehr, main = "east  horizon")
sum(is.na(values(ehr)))




#basins
basins <- rast("../spatial_data/NAD83UTM17N_epsg26917/GRASS_output/hpb_basin750.tif")
plot(basins, main = "basins")
sum(is.na(values(basins)))


# half basins
halfbasins <- rast("../spatial_data/NAD83UTM17N_epsg26917/GRASS_output/hpb_halfbasin750.tif")
plot(halfbasins, main = "halfbasins")
sum(is.na(values(halfbasins)))


#streams
streams <- rast("../spatial_data/NAD83UTM17N_epsg26917/GRASS_output/hpb_stream750.tif")
plot(streams, main = "streams", type = "classes")


```


## some plots to conceputalize basins 

```{r}
#basins and their streams
plot(basins)
plot(streams, type = "classes", add = T)
plot(streams, type = "classes", add = T,
          col  = c("steelblue", "orange", "forestgreen")) # one color per class

# col  = hcl.colors(n = 5, palette = "Set 2"),   # n = number of classes


#half basins
plot(halfbasins)



```



## fill NAs

function

```{r}
#function from chatt
fill_edge_na <- function(r, watershed, maxiter = 20) {
  
  # Ensure aligned
  r <- resample(r, watershed, method="near")
  
  # Mask to watershed
  r <- mask(r, watershed)
  
  for (i in 1:maxiter) {
    
    na_cells <- is.na(r) & !is.na(watershed)
    if (!any(values(na_cells), na.rm=TRUE)) break
    
    r <- focal(
      r,
      w = 3,
      fun = function(x, ...) {
        if (all(is.na(x))) return(NA)
        return(x[which(!is.na(x))[1]])
      },
      na.policy = "only",
      fillvalue = NA
    )
    
    r <- mask(r, watershed)
  }
  
  return(r)
}#end function

```


fill NAs for basin
```{r}

#basins
basins_filled  <- fill_edge_na(basins, ws)
plot(basins_filled)
sum(is.na(values(basins_filled)))
sum(is.na(values(basins)))
sum(is.na(values(ws)))


#basins
halfbasins_filled  <- fill_edge_na(halfbasins, ws)
plot(halfbasins_filled)
sum(is.na(values(halfbasins_filled)))
sum(is.na(values(halfbasins)))
sum(is.na(values(ws)))


##write new rasters
# writeRaster(basins_filled, "spatial_data/NAD83UTM17N_epsg26917/hpb_basin750_filled_15apr26.tif", overwrite = T)
# 
# writeRaster(halfbasins_filled, "spatial_data/NAD83UTM17N_epsg26917/hpb_halfbasin750_filled_15apr26.tif", overwrite = T)


```



####fix NLCD to watershed

```{r}
getwd()

#HPB watershed
ws <- rast("../spatial_data/NAD83UTM17N_epsg26917/hpb_ws.tif")
plot(ws, main = "Watershed")
sum(is.na(values(ws)))


#proejct NLCD to HPB WS
# nlcdOG <- rast("../../Downloaded_Data/NLCD/CCRbox_NLCD_Land_Cover_2021.tif")
# plot(nlcdOG)

nlcd <- rast("../../Downloaded_Data/NLCD/NLCD_rhessys_veg_cover.tif")
plot(nlcd)

NLCD_proj = project(as(nlcd, "SpatRaster"), ws, method = "near")
plot(NLCD_proj)

#now crop and mask
nlcd_crop <- crop(NLCD_proj, ws)
plot(nlcd_crop)

nlcd_mask <- mask(nlcd_crop, ws)
plot(nlcd_mask)
sum(is.na(values(nlcd_mask)))


##fill the NAs
nlcd_filled  <- fill_edge_na(nlcd_mask, ws)
plot(nlcd_filled)
sum(is.na(values(nlcd_filled)))
sum(is.na(values(nlcd_mask)))
sum(is.na(values(ws)))

# writeRaster(nlcd_filled, "../spatial_data/NAD83UTM17N_epsg26917/hpb_NLCD.tif", overwrite = T)


```



## try making a more informed patch map 
kmeans patches
same size patches across hillslopes

```{r}
##read in data sets of Dem and hillslopes
plot(dem)
sum(is.na(values(dem))) #number of non na cells
terra::expanse(dem, unit = "km") #area of watershed in km2

hillslopes <- halfbasins_filled
plot(hillslopes)
sum(is.na(values(hillslopes)))

### Extract non-NA cells with their coordinates and elevation
xy  <- as.data.frame(dem, xy = TRUE, na.rm = TRUE)

test <- terra::extract(hillslopes, as.matrix(xy[, c("x","y")]))
hsl <- test[, 1]  # or test$WHATEVER_THE_COLUMN_IS_CALLED

xy$hillslope <- hsl

names(xy)


n_total <- 300  # target total patches

# Count non-NA cells per hillslope
hs_vals   <- unique(na.omit(values(hillslopes)))
hs_counts <- sapply(hs_vals, function(h) sum(values(hillslopes) == h, na.rm = TRUE))
total_cells <- sum(hs_counts)

# Allocate patches proportional to hillslope area
hs_npatches <- round(hs_counts / total_cells * n_total)
hs_npatches <- pmax(hs_npatches, 1)  # at least 1 patch per hillslope

xy$patch_id <- NA
for (i in seq_along(hs_vals)) {
  hs  <- hs_vals[i]
  idx <- which(xy$hillslope == hs)
  sub <- xy[idx, c("x", "y", "NED_bathy@PERMANENT")]
  set.seed(42)
  km  <- kmeans(scale(sub), centers = hs_npatches[i], nstart = 5, iter.max = 100)
  xy$patch_id[idx] <- km$cluster + sum(hs_npatches[seq_len(i - 1)])
}

patch_map <- dem
patch_map[!is.na(dem)] <- xy$patch_id
patch_map <- mask(patch_map, hillslopes)



#### PLOT patch maps
# pal <- colorRampPalette(c("darkblue", "white", "darkred"))
# plot(patch_map, col = pal(200))
plot(patch_map)
hillslope_poly <- as.polygons(hillslopes)
plot(hillslope_poly, add = TRUE, border = "black", lwd = 2, col = NA)
global(patch_map, "max", na.rm = TRUE)


##check number of patches in smaller hillslopes
print("patches Hillslope 1")
xy |> filter(hillslope == 1) |> pull(patch_id) |> unique()

print("patches Hillslope 2")
xy |> filter(hillslope == 2) |> pull(patch_id) |> unique()

###write and check raster
writeRaster(patch_map, "./spatial_data/NAD83UTM17N_epsg26917/hpb_patch_kmeans300.tif", overwrite = TRUE)
 
# plot(rast("./spatial_data/NAD83UTM17N_epsg26917/hpb_patch_kmeans300.tif"))
# unique(rast("./spatial_data/NAD83UTM17N_epsg26917/hpb_patch_kmeans300.tif"))
# sum(is.na(values(rast("./spatial_data/NAD83UTM17N_epsg26917/hpb_patch_kmeans300.tif")))) #number of non na cells


```





#run preprocessing with smaller patch sizes

```{r}

getwd()
## set up file paths 
template = "templateNLCD_hpb.txt"
type = "raster"
typepars = "../spatial_data/NAD83UTM17N_epsg26917" #path of the folder containing your rasters; these are copied from grass output in "../Spatial_Data/GRASS_hpbNEW_outputs/"
name = "hpb_mixed"
overwrite = FALSE
streams = "hpb_stream750"  #this is the GRASS streams that has segment #s ; doens't seem to affect flow table or worldfie using stream with segments or all segments as 1
header = FALSE
parallel = TRUE
make_stream = 4

## run preprocess
RHESSysPreprocess(
  template,
  name,
  map_dir = typepars,
  typepars = NULL,
  streams = streams,
  overwrite = overwrite,
  roads = NULL,
  impervious = NULL,
  roofs = NULL,
  header = header,
  meta = FALSE,
  asprules = NULL,
  unique_strata_ID = TRUE,
  seq_patch_IDs = FALSE,
  output_patch_map = FALSE,
  fire_grid_out = FALSE,
  parallel = parallel,
  make_stream = make_stream,
  skip_hillslope_check = TRUE,
  convert_aspect = TRUE,
  wrapper = TRUE
)


```


## Look at how veg ID fell out within patches

Function to get veg ID from worldfile
```{r}
parse_worldfile <- function(filepath) {
  lines <- readLines(filepath)
  
  patches <- list()
  current_patch <- NULL
  in_patch <- FALSE
  
  for (line in lines) {
    line <- trimws(line)
    
    # Detect patch block
    if (grepl("^\\d+\\s+patch_ID", line)) {
      if (!is.null(current_patch)) {
        patches <- append(patches, list(current_patch))
      }
      current_patch <- list(patch_ID = as.integer(strsplit(line, "\\s+")[[1]][1]))
      in_patch <- TRUE
      
    } else if (in_patch) {
      if (grepl("veg_parm_ID", line)) {
        current_patch$veg_parm_ID <- as.integer(strsplit(line, "\\s+")[[1]][1])
      }
      # Stop collecting at stratum level
      if (grepl("^\\d+\\s+stratum_ID", line)) {
        in_patch <- FALSE
      }
    }
  }
  
  # Catch last patch
  if (!is.null(current_patch)) {
    patches <- append(patches, list(current_patch))
  }
  
  bind_rows(patches)
}

```


get info from worldfile and plot

```{r}
getwd()
#### Parse worldfile 
df <- parse_worldfile("./hpb_mixed.world")
head(df)


#### patch map ----

# Load patch raster
patch_rast <- rast("../spatial_data/NAD83UTM17N_epsg26917/hpb_patch_kmeans300.tif")

plot(patch_rast)

# Build lookup table
lookup <- setNames(df$veg_parm_ID, df$patch_ID)

# Remap patch IDs to veg_parm_ID using classify()
# Create a reclassification matrix: [from, to]
reclass_mat <- cbind(
  from = as.numeric(names(lookup)),
  to   = as.numeric(lookup)
)

veg_rast <- classify(patch_rast, reclass_mat, others = NA)
sum(!is.na(values(veg_rast)))

#plot
plot(veg_rast)



#proejct NLCD to CCR WS
nlcdOG <- rast("../../Downloaded_Data/NLCD/CCRbox_NLCD_Land_Cover_2021.tif")
# plot(nlcdOG)
NLCD_proj <- project(nlcdOG, "EPSG:26917")
NLCD_proj <- project(as(NLCD_proj, "SpatRaster"), patch_rast, method = "near") |> 
  crop(patch_rast) |>  mask(patch_rast)
plot(NLCD_proj)


nlcd <- rast("../spatial_data/NAD83UTM17N_epsg26917/hpb_NLCD.tif")
plot(nlcd)

```












