---
title: "CCR_preprocess"
output: html_document
date: "2026-04-20"
---


## Get packages 

```{r}
library(terra)
library(RHESSysPreprocessing)
```


##reproject DEM to utm for GRASS
In GRASS using this dem JOINED to the bathymetry "Downloaded_data/CCR_bathy/CCR_bathy_DEM_bind.tif" as input that was created in GETDEM.R
And the associated watershed file from that script "Downloaded_data/Watershed_rasters/ccr_ws.tif"

In GRASS making slope, aspect, basins, etc maps 
Will read these in and clean up below, then format veg cover maps




#### look at maps 
sum(is.na(values(ws))) is to check for NAs, if value doesn't equal basins will need to fill those gaps 

```{r}
#basin
getwd()
ws <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_ws.tif")
plot(ws, main = "Watershed")
sum(is.na(values(ws)))

#dem
dem <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_dem.tif")
plot(dem, main = "DEM")
sum(is.na(values(dem)))


#slope
slope <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_slope.tif")
plot(slope, main = "slope")
sum(is.na(values(slope)))


#aspect
aspect <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_aspect.tif")
plot(aspect, main = "aspect")
sum(is.na(values(aspect)))


#West Horizon
whr <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_whr.tif")
plot(whr, main = "West horizon")
sum(is.na(values(whr)))


#East horizon
ehr <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_ehr.tif")
plot(ehr, main = "east  horizon")
sum(is.na(values(ehr)))




#basins
basins1K <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_basin1K.tif")
plot(basins1K, main = "basins 1K")
sum(is.na(values(basins1K)))



# half basins
halfbasins1K <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_halfbasin1K.tif")
plot(halfbasins1K, main = "halfbasins 1K")
sum(is.na(values(halfbasins1K)))



#streams
streams1K <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_stream1K.tif")
plot(streams1K, main = "streams 1K", type = "classes")
unique(streams1K)



```


## some plots to conceputalize basins 

```{r}
#basins and their streams
plot(basins1K)
plot(streams1K, type = "classes", add = T)
plot(streams1K, type = "classes", add = T,
     col  = hcl.colors(n = 34, palette = "Set 2"),   # n = number of classes
     #col  = c("steelblue", "orange", "forestgreen") # one color per class
     ) 



```



## fill NAs

function

```{r}
#function from chatt
fill_edge_na <- function(r, watershed, maxiter = 20) {
  
  # Ensure aligned
  r <- resample(r, watershed, method="near")
  
  # Mask to watershed
  r <- mask(r, watershed)
  
  for (i in 1:maxiter) {
    
    na_cells <- is.na(r) & !is.na(watershed)
    if (!any(values(na_cells), na.rm=TRUE)) break
    
    r <- focal(
      r,
      w = 3,
      fun = function(x, ...) {
        if (all(is.na(x))) return(NA)
        return(x[which(!is.na(x))[1]])
      },
      na.policy = "only",
      fillvalue = NA
    )
    
    r <- mask(r, watershed)
  }
  
  return(r)
}#end function

```


fill NAs for basin
```{r}

#basins
basins1K_filled  <- fill_edge_na(basins1K, ws)
plot(basins1K_filled)
sum(is.na(values(basins1K_filled)))
sum(is.na(values(basins1K)))
sum(is.na(values(ws)))

# writeRaster(basins1K_filled, "../spatial_data/ccr_basin1K_filled.tif", overwrite = T)


#half basins
halfbasins1K_filled  <- fill_edge_na(halfbasins1K, ws)
plot(halfbasins1K_filled)
sum(is.na(values(halfbasins1K_filled)))
sum(is.na(values(halfbasins1K)))
sum(is.na(values(ws)))

# writeRaster(halfbasins1K_filled, "../spatial_data/ccr_halfbasin1K_filled.tif", overwrite = T)


```


####fix NLCD to watershed

```{r}
getwd()

#CCR watershed
ws <- rast("../GRASSccr_ned_dem_OUTPUTS/ccr_ws.tif")
plot(ws, main = "Watershed")
sum(is.na(values(ws)))


#proejct NLCD to CCR WS
# nlcdOG <- rast("../../Downloaded_Data/NLCD/CCRbox_NLCD_Land_Cover_2021.tif")
# plot(nlcdOG)

nlcd <- rast("../../Downloaded_Data/NLCD/NLCD_rhessys_veg_cover.tif")
plot(nlcd)

NLCD_proj = project(as(nlcd, "SpatRaster"), ws, method = "near")
plot(NLCD_proj)

#now crop and mask
nlcd_crop <- crop(NLCD_proj, ws)
plot(nlcd_crop)

nlcd_mask <- mask(nlcd_crop, ws)
plot(nlcd_mask)
sum(is.na(values(nlcd_mask)))


##fill the NAs
nlcd_filled  <- fill_edge_na(nlcd_mask, ws)
plot(nlcd_filled)
sum(is.na(values(nlcd_filled)))
sum(is.na(values(nlcd_mask)))
sum(is.na(values(ws)))

# writeRaster(nlcd_filled, "../spatial_data/ccr_NLCD.tif", overwrite = T)


```


## Fix reservoir to basins and stream numbers
This a mess; skipping for now and just using basins from above 

```{r}
# 
# # --- Load inputs ---
# watershed_r <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_ws.tif")
# subbasins_r <- rast("../spatial_data/ccr_basin1K_filled.tif")
# streams_r   <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_stream1K.tif")
# dem_r       <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_dem.tif")
# 
# # ── Step 1: Create reservoir mask ────────────────────────────────────────────
# # All cells at or below 356.6m become the reservoir subbasin (ID = 9999)
# reservoir_mask <- ifel(dem_r <= 356.7, 1L, NA)
# 
# message("Reservoir cells: ", global(reservoir_mask, "sum", na.rm = TRUE)[1, 1])
# 
# # ── Step 2: Update subbasin map ──────────────────────────────────────────────
# # Any cell that is reservoir gets ID 9999, all others keep their original ID
# subbasins_updated <- ifel(!is.na(reservoir_mask), 9999L, subbasins_r)
# 
# message("Subbasin IDs (should include 9999): ")
# print(sort(unique(na.omit(values(subbasins_updated)))))
# 
# # ── Step 3: Update stream map ─────────────────────────────────────────────────
# # Each stream cell inherits the subbasin ID it spatially falls within
# streams_updated <- ifel(!is.na(streams_r), subbasins_updated, NA)
# 
# # ── Step 4: Verify every subbasin has at least one stream cell ───────────────
# basin_ids   <- sort(unique(na.omit(values(subbasins_updated))))
# stream_ids  <- sort(unique(na.omit(values(streams_updated))))
# missing_str <- setdiff(basin_ids, stream_ids)
# 
# message("\nBasins with no stream cell: ",
#         ifelse(length(missing_str) == 0, "none",
#                paste(missing_str, collapse = ", ")))
# 
# # # ── Step 5: For basins missing a stream, inject one at the lowest DEM cell ───
# # if (length(missing_str) > 0) {
# #   for (bid in missing_str) {
# #     basin_cells  <- which(values(subbasins_updated) == bid)
# #     outlet_cell  <- basin_cells[which.min(values(dem_r)[basin_cells])]
# #     values(streams_updated)[outlet_cell] <- bid
# #     message("  Injected stream cell for basin ", bid,
# #             " at cell ", outlet_cell,
# #             " (elev = ", round(values(dem_r)[outlet_cell], 2), "m)")
# #   }
# # }
# 
# # ── Step 6: Final checks ──────────────────────────────────────────────────────
# basin_ids_final  <- sort(unique(na.omit(values(subbasins_updated))))
# stream_ids_final <- sort(unique(na.omit(values(streams_updated))))
# still_missing    <- setdiff(basin_ids_final, stream_ids_final)
# 
# message("\n--- Summary ---")
# message("Total subbasins : ", length(basin_ids_final))
# message("Total stream IDs: ", length(stream_ids_final))
# message("Basins still missing streams: ",
#         ifelse(length(still_missing) == 0, "none",
#                paste(still_missing, collapse = ", ")))
# 
# message("\nSubbasin cell counts:")
# print(freq(subbasins_updated))
# 
# message("\nStream cell counts per basin:")
# print(freq(streams_updated))
# 
# ##plot 
# plot(streams_updated, type = "classes")
# plot(subbasins_updated, type = "classes")

```







## try making a more informed patch map 
kmeans patches
same size patches across hillslopes


```{r}
# ── Patch map generation ──────────────────────────────────────────────────────
library(terra)

dem        <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_dem.tif")
hillslopes <- rast("../spatial_data/ccr_halfbasin1K_filled.tif")

elev_col <- names(dem)  # grabs whatever the DEM band is called

# ── Extract non-NA cells ──────────────────────────────────────────────────────
xy   <- as.data.frame(dem, xy = TRUE, na.rm = TRUE)
test <- terra::extract(hillslopes, as.matrix(xy[, c("x", "y")]))
xy$hillslope <- test[, 1]

# ── Diagnose flat hillslopes before clustering ────────────────────────────────
hs_vals   <- unique(na.omit(values(hillslopes)))
hs_counts <- sapply(hs_vals, function(h) sum(values(hillslopes) == h, na.rm = TRUE))

flat_hs <- hs_vals[sapply(hs_vals, function(h) {
  elev <- xy[xy$hillslope == h, elev_col]
  sd(elev, na.rm = TRUE) == 0
})]
message("Flat hillslopes (zero elevation variance): ",
        ifelse(length(flat_hs) == 0, "none", paste(flat_hs, collapse = ", ")))

# ── Allocate patches proportional to hillslope area ──────────────────────────
# Exclude flat hillslopes from proportional allocation — they get 1 patch each
n_total      <- 1000
land_hs      <- hs_vals[!hs_vals %in% flat_hs]
land_counts  <- hs_counts[!hs_vals %in% flat_hs]
total_cells  <- sum(land_counts)

hs_npatches_land <- round(land_counts / total_cells * n_total)
hs_npatches_land <- pmax(hs_npatches_land, 1)

# ── Assign flat hillslopes a single patch ID each ────────────────────────────
# These are reservoir/flat areas — no point clustering them
flat_patch_start <- sum(hs_npatches_land) + 1

xy$patch_id <- NA

for (j in seq_along(flat_hs)) {
  fhs <- flat_hs[j]
  idx <- which(xy$hillslope == fhs)
  xy$patch_id[idx] <- flat_patch_start + (j - 1)
  message("Flat hillslope ", fhs, " → single patch ID ", flat_patch_start + (j - 1))
}

# ── Cluster land hillslopes ───────────────────────────────────────────────────
for (i in seq_along(land_hs)) {
  hs  <- land_hs[i]
  idx <- which(xy$hillslope == hs)
  sub <- xy[idx, c("x", "y", elev_col)]
  
  sub <- sub[complete.cases(sub) & apply(sub, 1, function(r) all(is.finite(r))), ]
  
  if (nrow(sub) == 0) {
    message("Hillslope ", hs, " has no valid cells — skipping")
    next
  }
  
  n_centers <- min(hs_npatches_land[i], nrow(sub))
  
  # Scale then zero out any column that produced non-finite values
  sub_scaled <- scale(sub)
  bad_cols   <- apply(sub_scaled, 2, function(col) any(!is.finite(col)))
  if (any(bad_cols)) {
    message("Hillslope ", hs, " has zero-variance column(s): ",
            paste(names(sub)[bad_cols], collapse = ", "), " — zeroing out")
    sub_scaled[, bad_cols] <- 0
  }
  
  set.seed(42)
  km <- kmeans(sub_scaled, centers = n_centers, nstart = 5, iter.max = 100)
  xy$patch_id[idx] <- km$cluster + sum(hs_npatches_land[seq_len(i - 1)])
  
  if (i %% 10 == 0) message("Processed ", i, " of ", length(land_hs), " hillslopes...")
}

# ── Check coverage ────────────────────────────────────────────────────────────
message("\nTotal patches created: ", length(unique(na.omit(xy$patch_id))))
message("Unassigned cells: ", sum(is.na(xy$patch_id)))

# ── Write patch raster ────────────────────────────────────────────────────────
patch_r <- rast(dem)
values(patch_r) <- NA

cell_idx <- cellFromXY(dem, as.matrix(xy[, c("x", "y")]))
values(patch_r)[cell_idx] <- xy$patch_id

# ── Plot ──────────────────────────────────────────────────────────────────────
plot(patch_r, type = "classes", main = "Patch map")
unique(patch_r)

#write patch map
# writeRaster(patch_r, "../spatial_data/ccr_patch_map_kmeans1000.tif", overwrite = TRUE)


```





#run preprocessing with smaller patch sizes

```{r}

getwd()
## set up file paths 
template = "templateNLCD_ccr.txt"
type = "raster"
typepars = "../spatial_data" #path of the folder containing your rasters
name = "ccr_mixed"
overwrite = FALSE
streams = "ccr_stream1K"  #this is the GRASS streams that has segment #s ; doens't seem to affect flow table or worldfie using stream with segments or all segments as 1
header = FALSE
parallel = TRUE
make_stream = 4

## run preprocess
RHESSysPreprocess(
  template,
  name,
  map_dir = typepars,
  typepars = NULL,
  streams = streams,
  overwrite = overwrite,
  roads = NULL,
  impervious = NULL,
  roofs = NULL,
  header = header,
  meta = FALSE,
  asprules = NULL,
  unique_strata_ID = TRUE,
  seq_patch_IDs = FALSE,
  output_patch_map = FALSE,
  fire_grid_out = FALSE,
  parallel = parallel,
  make_stream = make_stream,
  skip_hillslope_check = TRUE,
  convert_aspect = TRUE,
  wrapper = TRUE
)


```















