---
title: "CCR_preprocess"
output: html_document
date: "2026-04-20"
---


## Get packages 

```{r}
library(terra)
library(tidyverse)
library(RHESSysPreprocessing)
```


##reproject DEM to utm for GRASS
In GRASS using the DEM where CCR is assumed to be at full pond, this is functinally the downloaded DEM but fixing weird affect near dam. I set any value <= 356.6 elevation to 356.6 "CCR_files/CCR_nhdburn/spatial_data/CCR_DEM_fullpond.tif" as input that was created in DEM_NHDburn.R
And the associated watershed file from that script "Downloaded_data/Watershed_rasters/ccr_ws.tif"

In GRASS making slope, aspect, basins, etc maps 
Will read these in and clean up below, then format veg cover maps




#### look at maps 
sum(is.na(values(ws))) is to check for NAs, if value doesn't equal basins will need to fill those gaps 

```{r}
#basin
getwd()
ws <- rast("./spatial_data/ccr_ws.tif")
plot(ws, main = "Watershed")
sum(is.na(values(ws)))

#dem
dem <- rast("./spatial_data/ccr_dem.tif")
plot(dem, main = "DEM")
sum(is.na(values(dem)))


#slope
slope <- rast("../spatial_data/ccr_slope.tif")
plot(slope, main = "slope")
sum(is.na(values(slope)))


#aspect
aspect <- rast("./spatial_data/ccr_aspect.tif")
plot(aspect, main = "aspect")
sum(is.na(values(aspect)))


#West Horizon
whr <- rast("./spatial_data/ccr_whr.tif")
plot(whr, main = "West horizon")
sum(is.na(values(whr)))


#East horizon
ehr <- rast("./spatial_data/ccr_ehr.tif")
plot(ehr, main = "east  horizon")
sum(is.na(values(ehr)))




#basins
basins1K <- rast("./spatial_data/ccr_basin1K.tif")
plot(basins1K, main = "basins 1K")
sum(is.na(values(basins1K)))

# half basins
halfbasins1K <- rast("./spatial_data/ccr_halfbasin1K.tif")
plot(halfbasins1K, main = "halfbasins 1K")
sum(is.na(values(halfbasins1K)))



#streams
streams1K <- rast("./spatial_data/ccr_stream1K.tif")
plot(streams1K, main = "streams 1K", type = "classes")
unique(streams1K)

# flow acc; may want for stream routing
flowacc <- rast("./spatial_data/flow_acc.tif")
plot(flowacc, main = "flow acc")
sum(is.na(values(flowacc)))


```


## some plots to conceputalize basins 

Basins with labels 
```{r}
#basins with labels
# Get centroids of each basin polygon
basin_polys <- as.polygons(basins1K)
basin_cents <- centroids(basin_polys)

# Extract coordinates and basin ID values
cents_df <- as.data.frame(basin_cents, geom = "XY")

# Plot the raster then overlay labels
plot(basins1K, type = "classes", legend = FALSE)
text(basin_cents, labels = basin_cents$`basin1K@PERMANENT`, cex = 0.6)
```


Basins and streams
```{r}
#basins and their streams
plot(basins1K)
plot(streams1K, type = "classes", add = T)
plot(streams1K, type = "classes", add = T,
     col  = hcl.colors(n = 33, palette = "Set 2"),   # n = number of classes
     #col  = c("steelblue", "orange", "forestgreen") # one color per class
     ) 



```



## fill NAs

function

```{r}
#function from chatt
fill_edge_na <- function(r, watershed, maxiter = 20) {
  
  # Ensure aligned
  r <- resample(r, watershed, method="near")
  
  # Mask to watershed
  r <- mask(r, watershed)
  
  for (i in 1:maxiter) {
    
    na_cells <- is.na(r) & !is.na(watershed)
    if (!any(values(na_cells), na.rm=TRUE)) break
    
    r <- focal(
      r,
      w = 3,
      fun = function(x, ...) {
        if (all(is.na(x))) return(NA)
        return(x[which(!is.na(x))[1]])
      },
      na.policy = "only",
      fillvalue = NA
    )
    
    r <- mask(r, watershed)
  }
  
  return(r)
}#end function

```


fill NAs for basin
```{r}

#basins
basins1K_filled  <- fill_edge_na(basins1K, ws)
plot(basins1K)
plot(basins1K_filled)
sum(is.na(values(basins1K_filled)))
sum(is.na(values(basins1K)))
sum(is.na(values(ws)))

# writeRaster(basins1K_filled, "./spatial_data/ccr_basin1K_filled.tif", overwrite = T)


#half basins
halfbasins1K_filled  <- fill_edge_na(halfbasins1K, ws)
plot(halfbasins1K)
plot(halfbasins1K_filled)
sum(is.na(values(halfbasins1K_filled)))
sum(is.na(values(halfbasins1K)))
sum(is.na(values(ws)))

# writeRaster(halfbasins1K_filled, "./spatial_data/ccr_halfbasin1K_filled.tif", overwrite = T)


```


####fix NLCD to watershed

```{r}
getwd()

#CCR watershed
ws <- rast("./spatial_data/ccr_ws.tif")
plot(ws, main = "Watershed")
sum(is.na(values(ws)))


#proejct NLCD to CCR WS
# nlcdOG <- rast("../../Downloaded_Data/NLCD/CCRbox_NLCD_Land_Cover_2021.tif")
# plot(nlcdOG)

nlcd <- rast("../../Downloaded_Data/NLCD/NLCD_rhessys_veg_cover_mixed.tif")
plot(nlcd)
#currnet nlcd key: 1= non veg, 2 = pasture, 3 = decid, 4 = evergreen, 5 = mixed forest

#change 2 to mixed for now
nlcd <- ifel(nlcd == 2, 5, nlcd)
plot(nlcd)

NLCD_proj = project(as(nlcd, "SpatRaster"), ws, method = "near")
plot(NLCD_proj)

#now crop and mask
nlcd_mask <- NLCD_proj |>  crop(ws) |> mask(ws)
plot(nlcd_mask)
sum(is.na(values(nlcd_mask)))

# writeRaster(nlcd_mask, "./spatial_data/ccr_NLCD.tif", overwrite = T)


## make soil map based on NLCD just to make the reservoir non loam
soil <- ifel(nlcd_mask == 1, 1, 8)

plot(soil)
sum(is.na(values(soil)))


# writeRaster(soil, "./spatial_data/ccr_SOIL.tif", overwrite = T)

```


make two stratum maps

```{r}
# library(terra)
# nlcd <- rast("../spatial_data/ccr_NLCD.tif")
# plot(nlcd)
# nonveg_id <- 1  # check your veg_nonveg.def stratum_default_ID and update this
# 
# # Primary stratum: mixed (5) → deciduous (3)
# veg1 <- classify(nlcd, cbind(5, 3))
# plot(veg1)
# writeRaster(veg1, "../spatial_data/ccr_veg_primary.tif", overwrite=TRUE)
# 
# # Secondary stratum: mixed (5) → evergreen (4), all others → nonveg
# veg2 <- classify(nlcd, rbind(c(5, 4)), others=nonveg_id)
# plot(veg2)
# writeRaster(veg2, "../spatial_data/ccr_veg_secondary.tif", overwrite=TRUE)
# 
# # Cover fractions
# cov1 <- classify(nlcd, cbind(5, 0.9), others=1.0)
# plot(cov1)
# writeRaster(cov1, "../spatial_data/ccr_cover_primary.tif", overwrite=TRUE)
# 
# cov2 <- classify(nlcd, cbind(5, 0.1), others=0.0)
# plot(cov2)
# writeRaster(cov2, "../spatial_data/ccr_cover_secondary.tif", overwrite=TRUE)

```



## try making a more informed patch map 
RES AREA GUIDED 
500 patches now 

```{r}
getwd()
# ── Settings ──────────────────────────────────────────────────────────────────

target_patches      <- 500
reservoir_elev      <- 356.7
res_split_threshold <- 80      # reservoir patches > this size get split
res_max_subpatches  <- 3
elev_weight         <- 1.5     # elevation influence in non-reservoir k-means

# ── Load and align ────────────────────────────────────────────────────────────

dem        <- rast("./spatial_data/ccr_dem.tif")
hillslopes <- rast("./spatial_data/ccr_halfbasin1K_filled.tif")
nlcd       <- rast("../../Downloaded_Data/NLCD/NLCD_rhessys_veg_cover_mixed.tif")

hillslopes <- resample(hillslopes, dem, method = "near")
hillslopes <- mask(hillslopes, dem)
hillslopes <- as.int(hillslopes)

xy <- xyFromCell(dem, seq_len(ncell(dem)))
df <- data.frame(
  cell = seq_len(ncell(dem)),
  x    = xy[, 1],
  y    = xy[, 2],
  dem  = values(dem,        mat = FALSE),
  hill = values(hillslopes, mat = FALSE)
)
df <- df[!is.na(df$dem) & !is.na(df$hill), ]
df$is_res <- df$dem <= reservoir_elev

message("Total valid cells: ",   nrow(df))
message("Reservoir cells: ",     sum(df$is_res))
message("Hillslopes: ",          length(unique(df$hill)))

df$patch <- NA_integer_

# =============================================================================
# STAGE 1 — Reservoir patches
# =============================================================================

message("\n--- Stage 1: Reservoir patches ---")

res_hills <- sort(unique(df$hill[df$is_res]))
next_id   <- 0L

for (h in res_hills) {
  rows    <- which(df$hill == h & df$is_res)
  n_cells <- length(rows)
  
  if (n_cells <= res_split_threshold) {
    next_id <- next_id + 1L
    df$patch[rows] <- next_id
  } else {
    k <- min(res_max_subpatches, max(2, floor(n_cells / res_split_threshold)))
    coords <- df[rows, c("x", "y")]
    set.seed(1)
    km <- kmeans(coords, centers = k, nstart = 5, iter.max = 50)
    df$patch[rows] <- next_id + km$cluster
    next_id <- next_id + k
    message("  Hillslope ", h, ": ", n_cells, " reservoir cells -> ", k, " patches")
  }
}

n_res <- next_id
message("Total reservoir patches: ", n_res)

# =============================================================================
# STAGE 2 — Allocate remaining budget
# =============================================================================

remaining_budget <- target_patches - n_res
message("\n--- Stage 2: Allocating ", remaining_budget, " non-reservoir patches ---")

nonres_counts <- table(df$hill[!df$is_res])
hill_ids_nr   <- as.integer(names(nonres_counts))
hill_areas    <- as.integer(nonres_counts)
total_nonres  <- sum(hill_areas)

hill_alloc <- round(remaining_budget * hill_areas / total_nonres)
hill_alloc[hill_alloc < 1] <- 1

while (sum(hill_alloc) > remaining_budget) {
  biggest <- which.max(hill_alloc)
  hill_alloc[biggest] <- hill_alloc[biggest] - 1
}
while (sum(hill_alloc) < remaining_budget) {
  underserved <- which.max(hill_areas / hill_alloc)
  hill_alloc[underserved] <- hill_alloc[underserved] + 1
}

names(hill_alloc) <- hill_ids_nr
message("Allocation summary across hillslopes:")
print(summary(hill_alloc))

# =============================================================================
# STAGE 3 — Non-reservoir k-means per hillslope
# =============================================================================

message("\n--- Stage 3: Elevation-weighted k-means per hillslope ---")

for (i in seq_along(hill_ids_nr)) {
  
  h       <- hill_ids_nr[i]
  n_allot <- hill_alloc[i]
  
  rows <- which(df$hill == h & !df$is_res)
  if (length(rows) == 0) next
  
  if (n_allot == 1 || length(rows) <= n_allot) {
    # Trivial: one patch (or fewer cells than allocated patches)
    next_id <- next_id + 1L
    df$patch[rows] <- next_id
    next
  }
  
  # Build feature matrix: standardized (x, y, elev * weight)
  # Standardization uses each variable's own sd within this hillslope
  xs <- df$x[rows]
  ys <- df$y[rows]
  es <- df$dem[rows]
  
  xs_s <- if (sd(xs) > 0) (xs - mean(xs)) / sd(xs) else xs * 0
  ys_s <- if (sd(ys) > 0) (ys - mean(ys)) / sd(ys) else ys * 0
  es_s <- if (sd(es) > 0) (es - mean(es)) / sd(es) else es * 0
  
  feats <- cbind(xs_s, ys_s, es_s * elev_weight)
  
  set.seed(i)  # reproducible per hillslope
  km <- tryCatch(
    kmeans(feats, centers = n_allot, nstart = 10, iter.max = 100),
    error = function(e) NULL
  )
  
  if (is.null(km)) {
    # Fall back: collapse to single patch if k-means fails
    next_id <- next_id + 1L
    df$patch[rows] <- next_id
    next
  }
  
  df$patch[rows] <- next_id + km$cluster
  next_id <- next_id + n_allot
}

# =============================================================================
# Compact
# =============================================================================

message("\n--- Finalizing ---")

unique_patches <- sort(unique(df$patch[!is.na(df$patch)]))
id_map <- setNames(seq_along(unique_patches), unique_patches)
df$patch <- id_map[as.character(df$patch)]

n_total <- max(df$patch, na.rm = TRUE)
message("Total patches: ", n_total, "  (target: ", target_patches, ")")

patch_map <- dem
values(patch_map) <- NA_integer_
patch_map[df$cell] <- df$patch
patch_map <- as.int(patch_map)

# =============================================================================
# Visualization
# =============================================================================
plot(patch_map)
unique(patch_map)

n <- max(values(patch_map), na.rm = TRUE)
set.seed(42)
plot(patch_map, col = sample(rainbow(n)), type = "classes",
     legend = FALSE, main = paste(n, "patches"))

# Draw patch outlines on top of the DEM
plot(dem, col = grey.colors(50), legend = FALSE, main = "Patch boundaries")
boundaries <- boundaries(patch_map, classes = TRUE)
plot(boundaries, add = TRUE, col = c(NA, "red"), legend = FALSE)

# Build a 3-class background: NA / reservoir / non-reservoir
bg <- dem
bg[] <- NA
bg[dem <= 356.7] <- 1   # reservoir
bg[dem >  356.7] <- 2   # non-reservoir

plot(bg,
     col = c("steelblue", "grey85"),
     type = "classes",
     levels = c("Reservoir (≤ 356.6 m)", "Upland"),
     main = "Reservoir extent + patch boundaries")

# Overlay patch boundaries on top
b <- boundaries(patch_map, classes = TRUE)
b[b == 0] <- NA   # drop interior cells so only edges plot
plot(b, add = TRUE, col = "red", legend = FALSE)

# =============================================================================
# Write raster
# =============================================================================

# writeRaster(patch_map, "./spatial_data/ccr_patch_map_500.tif", overwrite = TRUE)




```





KMEANS PATCHES
same size patches across hillslopes
This currnetly does 1500 but is labeled 1000

```{r}
# ── Patch map generation ──────────────────────────────────────────────────────
library(terra)

dem        <- rast("./spatial_data/ccr_dem.tif")
hillslopes <- rast("./spatial_data/ccr_halfbasin1K_filled.tif")

elev_col <- names(dem)  # grabs whatever the DEM band is called

# ── Extract non-NA cells ──────────────────────────────────────────────────────
xy   <- as.data.frame(dem, xy = TRUE, na.rm = TRUE)
test <- terra::extract(hillslopes, as.matrix(xy[, c("x", "y")]))
xy$hillslope <- test[, 1]

# ── Diagnose flat hillslopes before clustering ────────────────────────────────
hs_vals   <- unique(na.omit(values(hillslopes)))
hs_counts <- sapply(hs_vals, function(h) sum(values(hillslopes) == h, na.rm = TRUE))

flat_hs <- hs_vals[sapply(hs_vals, function(h) {
  elev <- xy[xy$hillslope == h, elev_col]
  sd(elev, na.rm = TRUE) == 0
})]
message("Flat hillslopes (zero elevation variance): ",
        ifelse(length(flat_hs) == 0, "none", paste(flat_hs, collapse = ", ")))

# ── Allocate patches proportional to hillslope area ──────────────────────────
# Exclude flat hillslopes from proportional allocation — they get 1 patch each
n_total      <- 1500
land_hs      <- hs_vals[!hs_vals %in% flat_hs]
land_counts  <- hs_counts[!hs_vals %in% flat_hs]
total_cells  <- sum(land_counts)

hs_npatches_land <- round(land_counts / total_cells * n_total)
hs_npatches_land <- pmax(hs_npatches_land, 1)

# ── Assign flat hillslopes a single patch ID each ────────────────────────────
# These are reservoir/flat areas — no point clustering them
flat_patch_start <- sum(hs_npatches_land) + 1

xy$patch_id <- NA

for (j in seq_along(flat_hs)) {
  fhs <- flat_hs[j]
  idx <- which(xy$hillslope == fhs)
  xy$patch_id[idx] <- flat_patch_start + (j - 1)
  message("Flat hillslope ", fhs, " → single patch ID ", flat_patch_start + (j - 1))
}

# ── Cluster land hillslopes ───────────────────────────────────────────────────
for (i in seq_along(land_hs)) {
  hs  <- land_hs[i]
  idx <- which(xy$hillslope == hs)
  sub <- xy[idx, c("x", "y", elev_col)]
  
  sub <- sub[complete.cases(sub) & apply(sub, 1, function(r) all(is.finite(r))), ]
  
  if (nrow(sub) == 0) {
    message("Hillslope ", hs, " has no valid cells — skipping")
    next
  }
  
  n_centers <- min(hs_npatches_land[i], nrow(sub))
  
  # Scale then zero out any column that produced non-finite values
  sub_scaled <- scale(sub)
  bad_cols   <- apply(sub_scaled, 2, function(col) any(!is.finite(col)))
  if (any(bad_cols)) {
    message("Hillslope ", hs, " has zero-variance column(s): ",
            paste(names(sub)[bad_cols], collapse = ", "), " — zeroing out")
    sub_scaled[, bad_cols] <- 0
  }
  
  set.seed(42)
  km <- kmeans(sub_scaled, centers = n_centers, nstart = 5, iter.max = 100)
  xy$patch_id[idx] <- km$cluster + sum(hs_npatches_land[seq_len(i - 1)])
  
  if (i %% 10 == 0) message("Processed ", i, " of ", length(land_hs), " hillslopes...")
}

# ── Check coverage ────────────────────────────────────────────────────────────
message("\nTotal patches created: ", length(unique(na.omit(xy$patch_id))))
message("Unassigned cells: ", sum(is.na(xy$patch_id)))

# ── Write patch raster ────────────────────────────────────────────────────────
patch_r <- rast(dem)
values(patch_r) <- NA

cell_idx <- cellFromXY(dem, as.matrix(xy[, c("x", "y")]))
values(patch_r)[cell_idx] <- xy$patch_id

# ── Plot ──────────────────────────────────────────────────────────────────────
plot(patch_r, type = "classes", main = "Patch map")
unique(patch_r)

#write patch map
# writeRaster(patch_r, "./spatial_data/ccr_patch_map_kmeans1000.tif", overwrite = TRUE)


```





#run preprocessing with smaller patch sizes

```{r}

getwd()
## set up file paths 
template = "template_ccr.txt"
type = "raster"
typepars = "./spatial_data" #path of the folder containing your rasters
name = "ccr_500"
overwrite = FALSE
streams = "ccr_stream1K"  #this is the GRASS streams that has segment #s ; doens't seem to affect flow table or worldfie using stream with segments or all segments as 1
header = FALSE
parallel = TRUE
make_stream = 4

## run preprocess
RHESSysPreprocess(
  template,
  name,
  map_dir = typepars,
  typepars = NULL,
  streams = streams,
  overwrite = overwrite,
  roads = NULL,
  impervious = NULL,
  roofs = NULL,
  header = header,
  meta = FALSE,
  asprules = NULL,
  unique_strata_ID = TRUE,
  seq_patch_IDs = FALSE,
  output_patch_map = FALSE,
  fire_grid_out = FALSE,
  parallel = parallel,
  make_stream = make_stream,
  skip_hillslope_check = TRUE,
  convert_aspect = TRUE,
  wrapper = TRUE
)


```



## Look at how veg ID fell out within patches

Function to get veg ID from worldfile
```{r}
parse_worldfile <- function(filepath) {
  lines <- readLines(filepath)
  
  patches <- list()
  current_patch <- NULL
  in_patch <- FALSE
  
  for (line in lines) {
    line <- trimws(line)
    
    if (grepl("^\\d+\\s+patch_ID", line)) {
      if (!is.null(current_patch)) {
        patches <- append(patches, list(current_patch))
      }
      current_patch <- list(patch_ID = as.integer(strsplit(line, "\\s+")[[1]][1]))
      in_patch <- TRUE
      
    } else if (in_patch) {
      if (grepl("veg_parm_ID", line)) {
        current_patch$veg_parm_ID <- as.integer(strsplit(line, "\\s+")[[1]][1])
      }
      if (grepl("soil_parm_ID", line)) {                                          # <-- add this
        current_patch$soil_parm_ID <- as.integer(strsplit(line, "\\s+")[[1]][1]) # <--
      }                                                                            # <--
      if (grepl("^\\d+\\s+stratum_ID", line)) {
        in_patch <- FALSE
      }
    }
  }
  
  if (!is.null(current_patch)) {
    patches <- append(patches, list(current_patch))
  }
  
  bind_rows(patches)
}

```


get info from worldfile and plot

```{r}
getwd()
#### Parse worldfile 
# df <- parse_worldfile("./ccr_mixed.world")
df <- parse_worldfile("./ccr_500.world")


head(df)

#### patch map ----

# Load patch raster
# patch_rast <- rast("./spatial_data/ccr_patch_map_kmeans1000.tif")
patch_rast <- rast("./spatial_data/ccr_patch_map_500.tif")

plot(patch_rast, main = "patch map")


stream_rast <- rast("./spatial_data/ccr_stream1K.tif")


# library(tmap)
# tm_shape(patch_rast)+
#   tm_raster(style = "cont", palette = "Spectral")

# --- veg_parm_ID raster (unchanged) ---
lookup_veg <- setNames(df$veg_parm_ID, df$patch_ID)
reclass_veg <- cbind(from = as.numeric(names(lookup_veg)), to = as.numeric(lookup_veg))
veg_rast <- classify(patch_rast, reclass_veg, others = NA)

# --- soil_parm_ID raster (new) ---
lookup_soil <- setNames(df$soil_parm_ID, df$patch_ID)
reclass_soil <- cbind(from = as.numeric(names(lookup_soil)), to = as.numeric(lookup_soil))
soil_rast <- classify(patch_rast, reclass_soil, others = NA)

# --- Plot both ---
plot(veg_rast,  main = "veg_parm_ID")
plot(soil_rast, main = "soil_parm_ID")

plot(soil_rast, main = "soil_parm_ID")
plot(stream_rast, add = T)

plot(ifel(veg_rast == 1, 1, 0))
plot(ifel(veg_rast == 3, 3, 0))


# #proejct NLCD to CCR WS
# nlcdOG <- rast("../../Downloaded_Data/NLCD/CCRbox_NLCD_Land_Cover_2021.tif")
# # plot(nlcdOG)
# NLCD_proj <- project(nlcdOG, "EPSG:26917")
# NLCD_proj <- project(as(NLCD_proj, "SpatRaster"), patch_rast, method = "near") |> 
#   crop(patch_rast) |>  mask(patch_rast)
# plot(NLCD_proj)
# 
# streams <- rast("../GRASSccr_dem_bathy_OUTPUTS/ccr_stream1K.tif")
# streams <- ifel(!is.na(streams), 1, NA)
# plot(streams, col = "black", add = T)
# 
# nlcd <- rast("../../Downloaded_Data/NLCD/NLCD_rhessys_veg_cover.tif")
# plot(nlcd)

```






